---
title: "BATS_sampling"
author: "Krista Longnecker"
date: "`28 May 2026`"
output: html_notebook
---


```{r load_libraries, message=FALSE, warning=FALSE}
rm(list = ls()) #start with a clean workspace (better for testing)

library(data.table) # you need to install this package from CRAN
library(dplyr)
library(readr) #this is part of the tidyverse package
library(here) #install this (once, install.packages("here))
#library(tidyverse) #amazing 
library(lubridate) #great for handling date data

# library(reshape2)
# library(MBA)
# library(mgcv)
# library(marmap)
# library(FNN)

library(readxl) #reading excel files
library(purrr)
#library(hms) #for time data
library(writexl)#Use the following to install castr if it doesnt load above.  See here for details https://rdrr.io/github/jiho/castr/
#install.packages("remotes")
#remotes::install_github("jiho/castr")
#library(castr)
```




```{r findDatafiles, message =FALSE}
baseDir = "D:/Dropbox/GitHub_niskin/data_pipeline/R_code"
# Define the path to the data file
wDir <- "../RawData/BATS & HS CAST SHEETS"
#three folders in there for BATS, BVal, and HS

batsDir <- "BATS Cast Sheets"
one <- paste0(wDir,"/",batsDir)

#sometimes there is a file labeled 'BIOS-SCOPE', other times there are just individual cast files that are relevant.
#start by pulling information from any file labeled 'BIOS-SCOPE'
patterns <- c("BIOS-SCOPE","H-R DOM") #one file with a different extension (cruise 10323)
possible <- paste(patterns,collapse="|")
files <- list.files(path = one,pattern = possible,recursive = TRUE, full.names = TRUE)


```

```{r doFirstFile}
#turns out BATS has a sample type that is listed as a number. Will be easier to 
#merge on that because the names change from sheet to sheet. I will need to confirm
#that I have guess right about the numbers.

#start with Luis's DNA. I think this is sample number eDNA (number 160)  
#use the first file to set up the dataframe...
idx <- 1
one <- read_excel(files[idx], skip=1,.name_repair = ~make.unique(.x, sep = "_")) #only 2+ instance is renamed
one <- one %>% mutate(across(everything(), as.character))

#remove last row (BATS repeats information)
one <- one[-nrow(one),]

#also delete the second Niskin column
one <- one[,-ncol(one)]

#add all cast information from the file...need some error checking
# (tidyup names...spelling and match to names/numbers for BIOS-SCOPE samples)

#definitely a better way to do this, but I am running out of steam
colnames(one) <- gsub("% cast","cast",colnames(one))
colnames(one) <- gsub("%cast","cast",colnames(one))
colnames(one) <- gsub("Nisken","niskin",colnames(one))
colnames(one) <- gsub("Niskin","niskin",colnames(one))

# BATS puts ## of sample type in different row, but that is the one that should match across cruises...
# search for the number and add to column names in main sheet if it does not already exist
k <- which (one[1,] > 0) 

# # keep the headers that correspond to the numbers, use this information to backtrack as needed
# one$sampleNames = colnames(one)[k]

# inefficient, but how my brain is thinking today
# iterate through all the columns with a number...make a column name if is does not exist
for (idx2 in 1:length(k)) {
  #colnames(one)[k][idx2] <- paste0(colnames(one)[k][idx2],"_",as.character(one[1,k[idx2]])) 
  colnames(one)[k][idx2] <- paste0(as.character(one[1,k[idx2]]) ,"_",colnames(one)[k][idx2]) 
}

df <- one
rm(one)


```

```{r doRestFiles}

#start with 2 because files[1] was used to setup the matrix

for (idx in 2:length(files)) {
  #idx <- 116
  one <- read_excel(files[idx], skip=1,.name_repair = ~make.unique(.x, sep = "_")) #only 2+ instance is renamed
  #need to make sure there are no empty column names...this happens when the excel file has empty columns...
  colnames(one)[colnames(one) == "" ] <- "empty"
  one <- one %>% mutate(across(everything(), as.character))
  
  #remove last row (BATS repeats information)
  one <- one[-nrow(one),]
  
  #add all cast information from the file...need some error checking
  # (tidyup names...spelling and match to names/numbers for BIOS-SCOPE samples)
  
  #definitely a better way to do this, but I am running out of steam
  colnames(one) <- gsub("% cast","cast",colnames(one))
  colnames(one) <- gsub("%cast","cast",colnames(one))
  colnames(one) <- gsub("Nisken","niskin",colnames(one))
  colnames(one) <- gsub("Niskin","niskin",colnames(one))

  # BATS puts ## of sample type in different row, but that is the one that should match across cruises...
  # search for the number and add to column names in main sheet if it does not already exist
  k <- which (one[1,] > 0) 
  
  # super inefficient, but how my brain is thinking today
  # iterate through all the columns with a number...make a column name if is does not exist
  for (idx2 in 1:length(k)) {
    #colnames(one)[k][idx2] <- paste0(colnames(one)[k][idx2],"_",as.character(one[1,k[idx2]])) 
    colnames(one)[k][idx2] <- paste0(as.character(one[1,k[idx2]]) ,"_",colnames(one)[k][idx2]) 
  }
  
  
  df <- bind_rows(df,one)
  rm(one)
  
}


```


```{r getUnique}


```


```{r}

write_xlsx(df,'../RawData/test8.xlsx')
```


MUSH below here, leave for now


```{r}

idx <- 62
one <- read_excel(files[idx], skip=1,.name_repair = ~make.unique(.x, sep = "_")) #only 2+ instance is renamed
#need to make sure there are no empty column names...this happens when the excel file has empty columns...
colnames(one)[colnames(one) == "" ] <- "empty"
one <- one %>% mutate(across(everything(), as.character))

#remove last row (BATS repeats information)
one <- one[-nrow(one),]

#add all cast information from the file...need some error checking
# (I know names changed as Niskin was spelled wrong for a while)
step1 = one[,1:2]

#definitely a better way to do this, but I am running out of steam
colnames(step1) <- gsub("% cast","cast",colnames(step1))
colnames(step1) <- gsub("%cast","cast",colnames(step1))
colnames(step1) <- gsub("Nisken","niskin",colnames(step1))
colnames(step1) <- gsub("Niskin","niskin",colnames(step1))

#find the sample type number in row 1:
k <- which (one[1,]==160)

t <- as.data.frame(one[,k])

step1[,"BS_eDNA_160"] <- t
#now, before I use the data, remove the first row
step1 <- step1[-1,]

df <- bind_rows(df,step1)

```





```{r}
#hard code the data frame
cols <- c("cast","niskin","depth","flag","BS_HDOM_148","BS_eDNA_160","BS_HRDOM_179","New_ID")
df <- data.frame(matrix(ncol = length(cols), nrow = 0))
colnames(df) <- cols
df <- df %>% mutate(across(everything(), as.character))

```


```{r tryNext, results='hide'}
#no idea how easy this will be to concatenate....let's try and see where the errors come up
#df <- map_df(files, readxl::read_excel) #this makes a mess, and does not merge based on columns (just appends rows)


# 2. Function to get headers without reading full data
get_headers <- function(file) {
  headers <- suppressMessages(names(read_excel(file, skip=1, n_max = 0)))
  return(headers)
}

# 3. Apply function to all files
all_headers <- lapply(files, get_headers)

#BATS sheets can have duplicate column names (e.g., two columns labeled 'BS_DOM')
#so R changes the name to BS_DOM...7 and BS_DOM...8 where the 6/7 are the column in the dataset

# # Optional: Name the list by filename for easy reference
names(all_headers) <- basename(files)
unique_headers <- sort(unique(tolower(unlist(all_headers))))

#looking at the list of headers the duplicates are going to make this ridiculous

```



```{r makeSquare2}
#try hard coding this...the automated is causing issues

df <- data.frame(matrix(ncol = length(unique_headers), nrow = 0))
colnames(df) = unique_headers
df <- df %>% mutate(across(everything(), as.character))

idx <- 1
one <- read_excel(files[idx], skip=1)
colnames(one) <- tolower(colnames(one))
one <- one %>% mutate(across(everything(), as.character))

df <- bind_rows(df,one)

idx <- 2
one <- read_excel(files[idx], skip=1)
colnames(one) <- tolower(colnames(one))
one <- one %>% mutate(across(everything(), as.character))

df <- bind_rows(df,one)

write_xlsx(df,'test5.xlsx')

```



```{r makeSquare}

#use the unique list of headers to make a square matrix...be sure to track 
#cruise/folder/year details so I can go back and figure out the issues

#make the data frame first:
df <- data.frame(matrix(ncol = length(unique_headers), nrow = 0))
colnames(df) = unique_headers

idx <- 1
one <- read_excel(files[idx], skip=1)
df <- bind_rows(list(df,one))

idx <- 2
one = as.data.frame(as.character(read_excel(files[idx], skip=1)))

df <- bind_rows(list(df,one))

```





``` {r}
library(data.table)

# colnames(one) <- tolower(colnames(one))
# rbindlist(list(df,one),use.names=TRUE,fill=TRUE)

df <- data.frame(matrix(ncol = length(unique_headers), nrow = 0))
colnames(df) = unique_headers

idx <- 1
one <- suppressMessages(read_excel(files[idx], skip=1))
colnames(one) <- tolower(colnames(one))

DT1 <- data.table(one)
DT2 <- data.table(df)

l = list(DT1,DT2)
rbindlist(l,fill=TRUE)


rbindlist(list(df,one),use.names=TRUE,fill=TRUE)



idx <- 2
one = suppressMessages(read_excel(files[idx], skip=1))
colnames(one) <- tolower(colnames(one))
df <- rbindlist(list(df,one),use.names=TRUE,fill=TRUE)

```

```{r}
library(tidyr)
library(readr)
result <- files <- list.files(path = one,pattern = "BIOS-SCOPE",recursive = TRUE, full.names = TRUE) %>% 
  map_df(~read_excel(.x) %>% mutate(across(.fns = as.character))) %>%
  type_convert()


```



# 2. Read and combine them into one data frame
# map_df applies read_excel to each file path and binds them by row
combined_data <- D %>% 
  map_df(~ read_excel(.))

# Assign names to the file vector so map_df can use them as IDs
names(files) <- basename(files)

combined_data <- files %>% 
  map_df(~ read_excel(.), .id = "source_file")


```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
